---
title: "Analysis of the ToothGrowth data in the R datasets package"
author: "Jason Shipp"
date: "Saturday, February 07, 2015"
output: html_document
---


### Introduction

The ToothGrowth data in the R datasets package contains data on the response length of teeth in each of 10 guinea pigs at each of three dose levels of Vitamin C (0.5, 1, or 2 mg)
with each of two delivery methods (orange juice or ascorbic acid).

The ToothGrowth data frame contains 60 observations of these 3 variables:

* [,1]	len: numeric; Tooth length in mm

* [,2]	supp: factor; Supplement type (VC or OJ)

* [,3]	dose: numeric; Dose in milligrams


### Data summary


Below is a summary of the unprocessed tooth data.

```{r kable, echo = TRUE, message = FALSE, warning = FALSE}
library(knitr)
ToothGrowth<- as.data.frame(ToothGrowth)
kable(summary(ToothGrowth))
```

Below is a summary of the unprocessed tooth data split by Supplement type; ascorbic acid (top) and orange juice (bottom).

```{r cache= FALSE}
library(knitr)
kable(summary(subset(ToothGrowth, supp=='VC')))
```

```{r echo = TRUE}
library(knitr)
kable(summary(subset(ToothGrowth, supp=='OJ')))
```

Below is a histogram of the tooth lengths.


```{r fig.width = 7, fig.height = 6, echo = FALSE}
hist(ToothGrowth$len, main = "Histogram of tooth lengths", xlab = "Frequency", ylab = "Tooth length in mm")
```

Below is a table summarising the tooth data by the mean tooth length per supplement and dose.

```{r echo = TRUE, message = FALSE, warning = FALSE}
library(doBy)
summaryBy(len ~ supp + dose, data = ToothGrowth, FUN = mean, keep.names=TRUE)
```

Below is a clustered bar graph representing the data in the above summary table.

```{r fig.width = 7, fig.height = 6, message = FALSE, warning = FALSE}
library(doBy)
ToothGrowth_summary<- summaryBy(len ~ supp + dose, data = ToothGrowth, FUN = mean, keep.names=TRUE)
library(ggplot2)
ggplot(ToothGrowth_summary, aes(x = dose, y= len)) + geom_bar(aes(fill = supp), position = "dodge", stat= 'identity') + labs(y = "Average tooth length in mm") + labs(x = "Supplement dose") +
ggtitle("Summary of ToothGrowth data") + theme(axis.title.x = element_text(size=20), axis.text.x = element_text(size=20)) + 
theme(axis.title.y = element_text(size=20), axis.text.y  = element_text(size=20)) + labs(fill = "Supplement") +
theme(legend.text = element_text(size=20)) + theme(legend.title = element_text(size=20)) 
```

The 0.9, 0.95 and 0.99 quantiles of the tooth lengths are calculated below.

```{r echo = TRUE}
quantile(ToothGrowth$len, c(0.9, 0.95, 0.99))
```

### Confidence intervals for the mean population tooth length for different supplements and doses

Assuming a **t-distribution**, the 95% confidence interval for the mean population tooth length is calculated below.

```{r echo = TRUE}
ToothGrowth_mean<- mean(ToothGrowth$len)
ToothGrowth_sd<- sd(ToothGrowth$len)
n<- length(ToothGrowth$len)
ToothGrowth_mean + (c(-1,1) * (qt(0.975, n-1)) * (ToothGrowth_sd/(sqrt(n))))
```

Assuming a **normal distribution**, the 95% confidence interval for the mean population tooth length is calculated below.

```{r echo = TRUE}
ToothGrowth_mean + (c(-1,1) * (qnorm(0.975)) * (ToothGrowth_sd/(sqrt(n))))
```

The above results of the 95% confidence interval using a normal distribution and t-distribution agree reasonably well with each other (to the nearest
whole number). We will therefore continue our analysis looking at differences in tooth growth by assuming a normal distribution.


Below are the 95% confidence intervals for the mean population tooth length for different doses of **ascorbic acid** (0.5, 1.0 and 2.0 milligrams).


```{r echo = TRUE}
x<- (subset(ToothGrowth, supp == 'VC' & dose == 0.5))$len
print(mean(x) + c(-1,1) * qnorm(0.975) * sd(x) / sqrt(length(x)))

x<- (subset(ToothGrowth, supp == 'VC' & dose == 1.0))$len
print(mean(x) + c(-1,1) * qnorm(0.975) * sd(x) / sqrt(length(x)))

x<- (subset(ToothGrowth, supp == 'VC' & dose == 2.0))$len
print(mean(x) + c(-1,1) * qnorm(0.975) * sd(x) / sqrt(length(x)))
```


Below are the 95% confidence intervals for the mean population tooth length for different doses of **orange juice** (0.5, 1.0 and 2.0 milligrams).

```{r echo = TRUE}
x<- (subset(ToothGrowth, supp == 'OJ' & dose == 0.5))$len
print(mean(x) + c(-1,1) * qnorm(0.975) * sd(x) / sqrt(length(x)))

x<- (subset(ToothGrowth, supp == 'OJ' & dose == 1.0))$len
print(mean(x) + c(-1,1) * qnorm(0.975) * sd(x) / sqrt(length(x)))

x<- (subset(ToothGrowth, supp == 'OJ' & dose == 2.0))$len
print(mean(x) + c(-1,1) * qnorm(0.975) * sd(x) / sqrt(length(x)))
```

The above results indicate that both increasing dosage and using orange juice in preference to ascorbic acid increase mean tooth length.
We can now use hypothesis tests to determine whether these differences in tooth length are significant.



### Hypothesis tests for determining whether the differences in mean tooth length for different supplements and doses are significant


The dose column of the ToothGrowth data frame has been simplified to two groups; "0-1" and "2", in order to make the column binary (only two groups).

```{r}
ToothGrowth[c("dose_simplified")]<- ifelse(ToothGrowth$dose <= 1, '0-1', '2')
```

T-tests have then been applied to the tooth lengths, to determine whether the lengths vary significantly by 1) supplement type and 2) supplement dosage. 

```{r echo = TRUE}
t.test(ToothGrowth$len ~ ToothGrowth$supp, paired=FALSE)
t.test(ToothGrowth$len ~ ToothGrowth$dose_simplified, paired=FALSE)
```

### Conclusions and Assumptions

The supplements t-test shows that the difference in mean tooth length between the two supplements is not quite 2 standard deviations, 
giving a p-value for the observed difference occurring by chance of just over 0.05; we therefore cannot reject the null hypothesis that the true difference 
in means is equal to 0. We therefore cannot be 95% certain that supplement type has a significant effect on tooth length.

The dosage t-test shows that the difference in mean tooth length between the two doses is over 2 standard deviations, 
giving a p-value for the observed difference occurring by chance of under 0.05; we therefore can reject the null hypothesis that the true difference 
in means is equal to 0. We therefore can be 95% certain that dosage has a significant positive effect on tooth length.

The assumptions for these t-tests are: 

* Tooth length shows a normal distribution

* Tooth length is a continuous variable

* Observations are independent of each other
