---
title: "Investigation of the Exponential Distribution in R"
author: "Jason Shipp"
date: "Saturday, February 07, 2015"
output: html_document
---


### Introduction


In this project, we will investigate the exponential distribution in R and compare it with the Central Limit Theorem. To achieve this, we will simulate 1000 groups of 40
random samples from the distribution, and take the mean of each of these simulations. We will then analyse the properties of the distribution of the means of these simulations.

The aims of the investigation are to:

* Show the simulated sample mean approximates the theoretical mean of the distribution.

* Show the simulated sample variance approximates the theoretical variance of the distribution.

* Show that the distribution is approximately normal.

The exponential distribution with rate lambda (λ) has density:
f(x) = λ {e}^{- λ x} for x ≥ 0.

The mean of the exponential distribution is 1/lambda, and the standard deviation is also 1/lambda. We will set lambda = 0.2 for all our simulations.

Below is a plot of the exponential distribution for 0<=x<=5:

```{r, echo= TRUE}
lambda<- 1
exp_dist_table<- as.data.frame(seq(from = 0, to = 5, by = 0.01))
names(exp_dist_table)[names(exp_dist_table)=="seq(from = 0, to = 5, by = 0.01)"] <- "x"
exp_dist_table[c("f(x)")]<- lambda  * (exp(-lambda * exp_dist_table$"x"))
plot(exp_dist_table, type = "l", lwd=2, col='red', main = "Exponential distribution curve")
```

### Methodology


The R function "rexp(n, lambda)" has been used to carry out 1000 simulations of the exponential distribution. Each simulation pulls a random sample of 40 
exponentials from the distribution. The mean and variance has been taken for each simulation row using the R functions "apply", "mean" and "var".
Assuming the simulation means are normally distributed, each mean has then been transformed to units of standard deviation with respect to the normal distribution using:
(mean - (population mean)) / (population standard deviation / sqrt(simulation samples)).

```{r, cache= FALSE}
n<- 40
lambda<- 0.2
pop_mean<- 1/lambda
pop_sd<- 1/lambda

simulations<- as.data.frame(as.numeric())
for(i in seq(1:1000))
{simulations<- rbind(simulations, c(rexp(n, lambda)))
}
```

```{r, echo= FALSE}
row.names(simulations)<- paste("simulation_", seq(1:length(simulations[,1])))
colnames(simulations)<- paste("n", seq(1:length(simulations[1,])), sep="")
simulations<- as.data.frame(simulations)
```

```{r, echo= TRUE}
simulations[c("simulation_mean")]<- apply(simulations, 1, mean)
simulations[c("standard_normal_transformation_of_mean")]<- with(simulations, ((simulation_mean - pop_mean)/(pop_sd/(sqrt(n)))))
simulations[c("simulation_variance")]<- apply(simulations, 1, var)
```

Below is the head of the resulting data frame, with the columns containing the simulations removed from view.


```{r kable, echo= FALSE}
library(knitr)
kable(head(simulations[, 41:43]), digits=3)
```

### Results


Below is a histogram of the 1000 simulation means; a black vertical line has been drawn through the hypothetical population mean (1/lambda = 5), 
and a red vertical line has been drawn through the mean of the simulated means. The bell shape of the distribution indicates it is normal.

```{r fig.width = 12, fig.height = 6, echo = FALSE}
hist(simulations$simulation_mean, main = "Histogram of exponential distribution simulation means", ylab = "Frequency", xlab = "Simulation mean") 
abline(v = (pop_mean))
abline(v = mean(simulations$simulation_mean), col = "red")
```

Below is a histogram of the transformed means, in units equivalent to the x-axis of a standard normal distribution. The histogram is overlain by a standard normal curve.
The bell-shaped distribution of the transformed simulation means indicates that the exponential distribution is normal. 


```{r fig.width = 7, fig.height = 6, echo = FALSE}
hist(simulations$standard_normal_transformation_of_mean, xlim=c(-4, 4), main = "Histogram of exponential distribution simulation means \n after transformation", ylab = "Frequency", xlab = "Transformed simulation mean")
par(new=TRUE)
x<- seq(from = -4, to = 4, by = 0.01 )
y<- dnorm(x, mean=0, sd=1)
plot(x,y,type="l", lwd=2,col="red", yaxt='n', xaxt='n', ann=FALSE)
```


Below is a density plot of the variances of the 1000 simulations; a black vertical line has been drawn through the hypothetical population variance ((1/lambda)^2 =25),
and a red vertical line has been drawn through the mean of the simulated variances.

```{r fig.width = 7, fig.height = 6, echo = FALSE}
plot(density(simulations$simulation_variance), main = "Histogram of exponential distribution simulation variances", ylab = "Density", xlab = "Simulation variance")
abline(v = (pop_sd) ^2)
abline(v = mean(simulations$simulation_variance), col="red")
```

### Conclusions

Below are the means of the simulated means and variances.

```{r echo= TRUE}
mean(simulations$simulation_mean)
mean(simulations$simulation_variance)
```

These compare well to the hypothetical population mean (1/lambda = 5) and variance ((1/lambda)^2 = 25). The errors are as follows.

Error in simulation mean:

```{r echo = FALSE}
paste(round(((5- mean(simulations$simulation_mean))/5)*100, 4), '%', sep = " ")
```

Error in simulation variance:

```{r echo = FALSE}
paste(round(((25- mean(simulations$simulation_variance))/25)*100, 4) , '%', sep = " ")
```

Below is the 95% confidence interval for the hypothetical population mean based on our simulations and the assumption of a normal distribution. 5 (the actual population mean)
Lies in this range.

```{r echo = TRUE}
mean(simulations$simulation_mean) + (c(-1,1) * (qnorm(0.975)) * (sd(simulations$simulation_mean)/(sqrt(length(simulations)))))
```

The 0.9, 0.95 and 0.99 quantiles for the transformed means are shown below, in order to determine if the distribution emulates a standard normal distribution.

```{r echo = TRUE}
quantile(simulations$standard_normal_transformation_of_mean, c(0.9, 0.95, 0.99))
```

The equivalent quantiles for the standard normal are shown below.

```{r echo = TRUE}
qnorm(c(0.9, 0.95, 0.99), mean = 0, sd = 1)
```

The closeness of these sets of quantiles, and the bell-shaped distribution of our simulation means, provide strong evidence for the exponential distribution being approximately normal.
